Generating Random Logic Programs Using Constraint Programming

### 2020-01-08 paper/paper.tex
Treating each clause uniformly, without dividing it into two parts
Bug fix in the code

\institute{University of Edinburgh, Edinburgh, UK\\
\email{p.dilkas@sms.ed.ac.uk}}
\maketitle

@@

\section{Heads of Clauses}

@@

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$, where $\Box$
    denotes a disabled clause,
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}$.
  \end{itemize}
\end{definition}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    i \ge \variable{arity} \implies \variable{arguments}[i] = |\constants{}| +
    |\variables{}| - 1.
  \]
\end{constraint}
In other words, we fix the unused arguments to be the last variable in
$\variables{}$ if there are any, and a constant otherwise. Because of the way
constants and variables are ordered, this is the only constant value in $\{ 0,
\dots, |\constants{}| + |\variables{}| - 1 \}$ that satisfies all other
constraints on the head.

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$, if $i \ge \variable{arity}$, then
  $\variable{arguments}[i] = \Box$, else $\variable{arguments}[i] \ne \Box$.
\end{constraint}

@@

\begin{constraint}
  For $i = 0, \dots, \maxNumClauses - 1$, let $p_i =
  \variable{headsOfClauses}[i].\variable{predicate}$. Then
  $(p_i)_{i=0}^{\maxNumClauses - 1}$ is sorted.
\end{constraint}

@@

\begin{constraint}
  Let $\prec$ be any total order defined over bodies of clauses, and let
  $\preceq$ be its extension with equality (in the same way as $\le$ extends
  $<$). If
  \[
    \variable{headsOfClauses}[i-1].\variable{predicate} =
    \variable{headsOfClauses}[i].\variable{predicate},
  \]
  then $\variable{bodiesOfClauses}[i-1] \preceq \variable{bodiesOfClauses}[i]$.
\end{constraint}
For example, $\preceq$ can be implemented as \variable{lexLessEq} over the
decision variables of each body.
% TODO: turn this into lexChainLessEq

\section{Counting Programs}
% TODO: update to account for the new definition of head

Let $p_a$ be the number of predicates in $\predicates{}$ with arity $a \in
\arities{}$.

Number of atoms (leaves of the tree):
\[
  A = \sum_{a \in \arities{}} p_a(|\variables{}| + |\constants{}|)^a
\]

Number of clauses:
\[
  C = 1 + \sum_{n=1}^{\maxNumNodes{}} T(n),
\]
where $T(n)$ is defined recursively as:
\[
  T(1) = A
\]
and
\[
  T(n) = T(n-1) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      k \ge 2,\\
      c_i \ge 1 \text{ for all } i}} \prod_{i=1}^k T(c_i).
\]

% TODO: explain how it works
Number of heads for a specific predicate with arity $a \in \arities{}$:
\[
  H_a = \sum_{v = 0}^a |\constants{}|^{a-v} \sum_{k=1}^{\min\{|\variables{}|,
    a\}} p_k(a),
\]
where $p_k(a)$ denotes the number of ways to partition integer $a$ into exactly
$k$ parts.

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{C}{h_i} H_{a_i}^{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n + k - 1}{k}.
\]

This is without any kind of cycle detection and without probabilities.
% TODO: describe what parameter values were used to test the model (or... not).

>>>>

\institute{University of Edinburgh, Edinburgh, UK\\
  Heriot-Watt University, Edinburgh, UK\\
  \email{p.dilkas@sms.ed.ac.uk}}
\maketitle

@@

\section{Heads of Clauses}
% TODO: section for clauses. define heads & bodies, then separate subsections
% for head and body constraints

@@

\begin{definition}
  The \emph{head} of a clause is composed of:
  \begin{itemize}
  \item a $\variable{predicate} \in \predicates \cup \{ \Box \}$.
  \item and $\arrayd{arguments}{\maxArity{}}{\constants{} \cup \variables{}}
    \cup \{ \Box \}$
  \end{itemize}
  In both cases, $\Box$ is the disabled value.
\end{definition}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
    \variable{arguments}[i] = \Box \iff i \ge \variable{arity}.
  \]
\end{constraint}

@@

\begin{constraint}
  For $i = 0, \dots, \maxArity{} - 1$,
  \[
  \variable{arguments}[i] = \Box \quad \iff \quad i \ge \variable{arity}.
  \]
\end{constraint}
% TODO: this could merged into a single constraint for termsPerClause

@@

@@

\begin{constraint}
  Clauses are sorted.
\end{constraint}

\section{Counting Programs}

This is without any kind of cycle detection and without probabilities.

Number of ways to fill $n$ positions with terms: % TODO: what's the smallest n?
\[
  P(n) = |\constants{}|^n + |\constants{}|^{n-1}|\variables{}| +
  \sum_{i=0}^{n-2} |\constants{}|^i \sum_{j=0}^{\min\{|\variables{}| - 1, n - i
    - 1\}} \binom{n-i-1}{j} (|\constants{}| + 1)^{n-i-j-1}
\]
where $i$ iterates over the number of constants before the first variable, and
$j$ iterates over the number of `new variable actions'. Explain:
$|\constants{}|+2$ possible actions.

Let $p_a$ be the number of predicates in $\predicates{}$ with arity $a \in
\arities{}$. Number of clauses with total arity $a$:
\begin{align*}
  C(0) &= 1,\\
  C(a) &= \sum_{n=1}^{\maxNumNodes{}} T(n, a),
\end{align*}
where $T(n, a)$ is defined recursively as:
\[
  T(1, a) = p_a
\]
and
\[
  T(n, a) = T(n-1, a) + 2\sum_{\substack{c_1 + \dots + c_k = n - 1,\\
      2 \le k \le \frac{a}{\min \arities{}},\\
      c_i \ge 1 \text{ for all } i}} \sum_{\substack{d_1 + \dots + d_k = a,\\
    d_i \ge \min \arities{} \text{ for all } i}} \prod_{i=1}^k T(c_i, d_i).
\]

Let us order the elements of $\predicates{}$, and let $a_i$ be the arity of the
$i$-th predicate. The number of programs is then:
\[
  \sum_{\substack{ \sum_{i=1}^{|\predicates{}|} h_i = n,\\
      |\predicates{}| \le n \le \maxNumClauses,\\
      h_i \ge 1 \text{ for all } i}} \prod_{i=1}^{|\predicates{}|}
  \multiset{\sum_{a=0}^{\maxArity{} \times \maxNumNodes{}} C(a) P(a+a_i)}{h_i},
\]
where
\[
  \multiset{n}{k} = \binom{n+k-1}{k}.
\]
